---
title: "6. Extracting amplitudes"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{6. Extracting amplitudes}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor: visual
editor_options:
  chunk_output_type: inline
---

> 🎯 **GOALS**
>
> *Extracting one value (mean ERP amplitude) per trial for statistical modeling.*

## 6.1 Read preprocessed epochs

-   Load packages and (artifact-corrected) epochs:

```{r, message=FALSE}
library(here)
library(eegUtils)
library(ggplot2)
```

```{r}
bids_dir <- here("data/n170")
deriv_dir <- here(bids_dir, "derivatives/eegUtils/sub-001/eeg")
preproc_file <- here(deriv_dir, "sub-001_task-N170_desc-preprocessed_eeg.rds")
epoch_file <- here(deriv_dir, "sub-001_task-N170_desc-corrected_eeg.rds")
dat_epo <- readRDS(epoch_file)
```

## 6.2 Select region of interest

-   Measures from a single EEG electrodes are noisy and spatially imprecise (e.g., volume conduction)

-   Averaging across nearby electrodes boosts signal-to-noise ratio

-   **Region of interest (ROI)** = the set of electrodes where we expect our ERP component of interest to be maximal

-   Convert epoched data to a single data frame for subsequent subsetting / averaging:

```{r}
dat_frame <- as.data.frame(dat_epo)
str(dat_frame)
```

-   Create a new "virtual" channel (i.e., column) containing the average signal in the ROI:

```{r}
roi <- c("P7", "P8", "P9", "P10", "PO7", "PO8")
dat_frame$roi <- rowMeans(dat_frame[roi])
```

## 6.2 Select time window

-   ERP components unfold on the time scale of 10 to multiple 100s of milliseconds

-   Select the **time window** where we expect our component to be (here: around 170 ms for the N170)

    -   Both region and time window of interest should be defined *before* looking at the data (i.e., preregistration)

```{r}
tmin <- 0.15
tmax <- 0.2
dat_frame <- subset(dat_frame, time >= tmin & time < tmax)
```

-   Averaging across this time window gets us one value (mean ERP amplitude) per trail

    -   Carry over the \`epoch_labels\` so we know the condition of each trial

```{r}
dat_trials <- aggregate(roi ~ epoch + epoch_labels, dat_frame, mean)
str(dat_trials)
```

-   A quick exploratory box plot of the single trial mean ERP amplitudes for faces vs. cars:

```{r}
ggplot(dat_trials, aes(x = epoch_labels, y = roi)) +
  geom_boxplot() +
  labs(x = NULL, y = "Mean amplitude (µV)") +
  theme_minimal()
```

-   We'll quantify this effect using statistical modeling in the next vignette

-   Save the single trial data for later:

```{r}
trials_file <- here(deriv_dir, "sub-001_task-N170_desc-trials_eeg.rds")
saveRDS(dat_trials, trials_file)
```

## Further reading

-   Luck, S. J., & Gaspelin, N. (2017). How to get statistically significant effects in any ERP experiment (and why you shouldn't). *Psychophysiology*, *54*(1), 146--157. <https://doi.org/10.1111/psyp.12639>

-   Paul, M., Govaart, G. H., & Schettino, A. (2021). Making ERP research more transparent: Guidelines for preregistration. *International Journal of Psychophysiology*, *164*, 52--63. <https://doi.org/10.1016/j.ijpsycho.2021.02.016>
