---
title: "6. Extracting amplitudes"
date: "2023-01-03"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{6. Extracting amplitudes}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor: visual
editor_options:
  chunk_output_type: inline
---

> 🎯 **GOALS**
>
> *Extracting one value (mean ERP amplitude) per trial for statistical modeling.*

## 6.1 Read preprocessed epochs

-   Load packages and (artifact-corrected) epochs:

```{r, message=FALSE}
library(here)
library(eegUtils)
library(ggplot2)
```

```{r}
bids_dir <- here("data/n170")
deriv_dir <- here(bids_dir, "derivatives/eegUtils/sub-001/eeg")
epoch_file <- here(deriv_dir, "sub-001_task-N170_desc-corrected_eeg.rds")
dat_epo <- readRDS(epoch_file)
```

## 6.2 Select region of interest

-   Measures from a single EEG electrodes are noisy and spatially imprecise (e.g., volume conduction)

-   Averaging across nearby electrodes boosts signal-to-noise ratio

-   **Region of interest (ROI)** = the set of electrodes where we expect our ERP component of interest to be maximal

-   First convert epoched data to a data frame for subsequent subsetting / averaging:

```{r}
dat_frame <- as.data.frame(dat_epo)
str(dat_frame)
```

-   Create a new "virtual" channel (i.e., column) containing the average signal in the ROI:

```{r}
roi <- c("P7", "P8", "P9", "P10", "PO7", "PO8")
dat_frame$roi <- rowMeans(dat_frame[roi])
```

## 6.2 Select time window

-   ERP components unfold on the time scale of 10 ms to multiple 100s of ms

-   Select the **time window** where we expect our component to be (here: around 170 ms for the N170)

    -   Note: Region and time window of interest should be defined *before* looking at the data (i.e., preregistration)!

```{r}
tmin <- 0.15
tmax <- 0.2
dat_frame <- subset(dat_frame, time >= tmin & time < tmax)
```

-   Averaging across this time window gets us one value (mean ERP amplitude) per trail

    -   Carry over the `epoch_labels` (coding the condition of each trial) and the `event_type` (coding the specific stimulus that was presented)

```{r}
dat_trials <- aggregate(
  roi ~ epoch + event_type + epoch_labels,
  dat_frame,
  mean
)
str(dat_trials)
```

```{r}
names(dat_trials) <- c("epoch", "stimulus", "condition", "amplitude")
```

-   Quick box plot comparing the single trial mean ERP amplitudes for faces vs. cars:

```{r}
ggplot(dat_trials, aes(x = condition, y = amplitude)) +
  geom_boxplot() +
  labs(x = NULL, y = "N170 amplitude (µV)") +
  theme_minimal()
```

-   We'll quantify this effect using statistical modeling in Vignette 7

-   Save the single trial data for later:

```{r}
trials_file <- here(deriv_dir, "sub-001_task-N170_trials.csv")
write.csv(dat_trials, trials_file, row.names = FALSE)
```

## Further reading

-   Luck, S. J., & Gaspelin, N. (2017). How to get statistically significant effects in any ERP experiment (and why you shouldn't). *Psychophysiology*, *54*(1), 146--157. <https://doi.org/10.1111/psyp.12639>

-   Paul, M., Govaart, G. H., & Schettino, A. (2021). Making ERP research more transparent: Guidelines for preregistration. *International Journal of Psychophysiology*, *164*, 52--63. <https://doi.org/10.1016/j.ijpsycho.2021.02.016>

## Add-on topics

### 6.3 Building a pipeline

-   Download the data from **all** participant (leaving out the `subject` argument in `get_erpcore()`)

```{r, message=FALSE, results="hide"}
erpcore::get_erpcore(
  component = "n170",
  dest_path = bids_dir,
  conflicts = "overwrite",
  type = "bids"
)
```

-   Define a **custom pipeline function**, combining all processing steps

    -   Input parameters: The EEG data (`.set`) file + all processing options

    -   Return value: A data frame with single trial mean ERP amplitudes (as above)

```{r}
run_pipeline <- function(set_file,
                         montage = "biosemi64",
                         reref_chans = "average",
                         reref_exclude = c("HEOG_left", "HEOG_right", "VEOG_lower"),
                         filter_low_freq = 0.1,
                         filter_high_freq = 30.0,
                         downsample_factor = 4.0,
                         event_codes = c(1:80),
                         epoch_labels = c(rep("face", 40), rep("car", 40)),
                         epoch_time_lim = c(-0.2, 0.8),
                         baseline = c(-0.2, 0.0),
                         artifact_threshold = c(-100.0, 100.0),
                         erp_roi = c("P7", "P8", "P9", "P10", "PO7", "PO8"),
                         erp_time_lim = c(0.15, 0.2)) {

  # From Vignette 3: Preprocessing
  dat <- import_set(set_file)
  dat <- electrode_locations(dat, montage = montage, overwrite = TRUE)
  dat <- eeg_reference(dat, reref_chans, reref_exclude)
  dat <- eeg_downsample(dat, downsample_factor)

  # From Vignette 4: Creating epochs
  dat <- epoch_data(dat, event_codes, epoch_time_lim, baseline, epoch_labels)
  dat <- ar_thresh(dat, artifact_threshold, reject = TRUE)

  # From Vignette 6: Extracting amplitudes
  dat <- as.data.frame(dat)
  dat$roi <- rowMeans(dat[roi])
  tmin <- min(erp_time_lim)
  tmax <- max(erp_time_lim)
  dat <- subset(dat, time >= tmin & time < tmax)
  dat <- aggregate(roi ~ epoch + event_type + epoch_labels, dat, mean)
  names(dat) <- c("epoch", "stimulus", "condition", "amplitude")
  
  return(dat)
}
```

-   Find all the `.set` EEG files inside our BIDS directory (including sub-directories):

```{r}
(set_files <- list.files(
  bids_dir,
  pattern = "\\.set$",
  recursive = TRUE,
  full.names = TRUE
))
```

-   Apply the pipeline function to each `.set` file, returning a list of data frames (one for each participant)

```{r, message=FALSE}
dats_group <- lapply(set_files, run_pipeline)
```

-   Combine all data frames into one big data frame:

    -   Adding an ID column to keep track which trials belong to which participant

```{r}
dat_group <- dplyr::bind_rows(dats_group, .id = "participant")
head(dat_group)
```

-   Save for later:

```{r}
group_file <- here(bids_dir, "derivatives/eegUtils/group_task-N170_trials.csv")
write.csv(dat_group, group_file, row.names = FALSE)
```
