---
title: "8. Making plots"
date: "2023-01-03"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{8. Making plots}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor: visual
editor_options:
  chunk_output_type: inline
---

> 🎯 **GOALS**
>
> *Visualizing ERP effects using time course plots and scalp topography plots.*

## 8.1 Time course

-   Load required packages:

    -   `ggplot2` for plotting

    -   `dplyr` and `tidyr` for data wrangling

```{r, message=FALSE}
library(here)
library(ggplot2)
library(dplyr)
library(tidyr)
library(eegUtils)
```

-   Load preprocessed epochs from a single ERP CORE participant:

```{r}
bids_dir <- here("data/n170")
deriv_dir <- here(bids_dir, "derivatives/eegUtils/sub-001/eeg")
epoch_file <- here(deriv_dir, "sub-001_task-N170_desc-corrected_eeg.rds")
dat_epo <- readRDS(epoch_file)
```

-   Create a **time course plot**:

    -   x-axis: Time

    -   y-axis: ERP amplitude (one electrode or ROI average)

    -   Colors: Average ERP wave forms in different conditions

```{r}
plot_timecourse(dat_epo, electrode = "PO7", colour = "epoch_labels")
```

-   Add annotations for the plotted time window and electrode of interest:

```{r}
tmin <- 0.15
tmax <- 0.2

plot_timecourse(dat_epo, electrode = "PO7", colour = "epoch_labels") +
  annotate("rect", xmin = tmin, xmax = tmax, ymin = -Inf, ymax = Inf, alpha = 0.2) +
  annotate("text", x = 0.7, y = 8, label = "Electrode PO8")
```

-   Plot all electrodes:

```{r}
erp_scalp(dat_epo, colour = "epoch_labels", size = 0.6)
```

## 8.2 Scalp topography

-   **Scalp topography** shows the distribution of voltages on the scalp

-   Either for a single condition or (more typically) for the difference between conditions

-   Requires some wrangling of the data:

    -   Goal: a data frame with a column of electrode names and a vector of amplitudes

    -   Using tidyverse style here (see the note in Vignette 1, Section 1.4)

```{r}
dat_epo %>%
  as.data.frame() %>%
  filter(time >= tmin & time < tmax) %>%
  select(-c(time, epoch, participant_id, recording, event_type)) %>%
  pivot_longer(-epoch_labels, names_to = "electrode", values_to = "amplitude") %>%
  group_by(electrode, epoch_labels) %>%
  summarize(amplitude = mean(amplitude), .groups = "drop") %>%
  group_split(epoch_labels) -> dats_topo

head(dats_topo[[1]])
head(dats_topo[[2]])
```

-   Create a new data frame that has the **difference** between face amplitudes and car amplitudes:

```{r}
dat_topo <- data.frame(
  electrode = dats_topo[[1]]$electrode,
  amplitude = dats_topo[[2]]$amplitude - dats_topo[[1]]$amplitude
)
head(dat_topo)
```

-   Create the topographic plot with `eegUtils`:

```{r}
topoplot(dat_topo, limits = c(-6.0, 6.0))
```
